---
title: "project"
output: html_document
---

### Introduction

This document outlines an exploratory data analysis of publicly available
assessment data for Idaho high schoolers. 

### Data


The dataset consists of 1,787 records, each representing a school in Idaho.
Within each record is 26 variables describing the following:

***[Identification]*** 

Variables that describe the school and the time of the survey (ncessch_num,
year, school_name, leaid_num, lea_name, fips, ncessch, leaid).

***[Demographics]***

Variables that describe the student population in question (grade_edfacts,
race, sex, lep, homeless, migrant, disability, foster_care, military_connected,
econ_disadvantaged).

***[Academics]***

Variables that describe the student population's academic performance
(read_test_num_valid, read_test_pct_prof_low, read_test_pct_prof_midpt,
read_test_pct_prof_high, math_test_num_valid, math_test_pct_prof_low,
math_test_pct_prof_midpt, math_test_pct_prof_high).


### EDA

***Math and Reading***

<Scatterplot of Math vs. Reading>

In the year 2017 there were 210 high schools that reported assessment results.


<Boxplot of Testing Scores>

Notably, reading scores trend higher than math scores. 


<Boxplot of Testing Scores (Gender)>

<Linear Regression (Gender)>

Additionally, reading scores differ fairly significantly by sex. A linear
regression shows that female students score roughly 12 points higher than their
male counterparts (with regards to midpoints).


<Boxplot of Testing Scores (Race)>

<Linear Regression (Race)>

Race represents a more complex set of relationships. 




### Methods and Objectives

***[Performance]*** 

*What factors are associated with high performance on the assessment? What
factors are associated with low performance?*




***[Prediction]*** 

*Can future assessment results be predicted with any level of reliability using
techniques covered in this course?* 

(Suggested methods: smoothing, random forest, boosting algorithms.)




***[Intervention]*** 

*Can we identify low performing schools and suggest a reasonable treatment
which might improve their assessment results?*

(Suggested methods: boxplots, clustering, review of domain-specific
literature.)




### Appendix A: Code


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(caret)
library(educationdata)
library(naniar)
library(rpart)
library(rpart.plot)
library(scales)
```

```{r helper_functions}
latest_year = 2017


fetch_idaho_school_data <- function(source, subtopic_name) {
  ## https://educationdata.urban.org/documentation/schools.html#edfacts-state-assessments-by-grade
  ##
  ## Note: "grade_edfacts = 9" translates to grades 9-12
  ## 
  ## Note: "fips = 16" translates to Idaho-only results
  ##
  ## Note: Idaho's state assessment switched providers in 2014, results should
  ##       reflect this change
  ##
  ## Edfacts
  if (source == 'edfacts') {
    if (length(subtopic_name) == 0) {
      data <- get_education_data(
        level = "schools",
        source = 'edfacts',
        topic = 'assessments',
        filters = list(grade_edfacts = 9, fips = 16, year = latest_year))
    }
    else {
      data <- get_education_data(
        level = "schools",
        source = 'edfacts',
        topic = 'assessments',
        filters = list(grade_edfacts = 9, fips = 16, year = latest_year),
        subtopic = subtopic_name)
    }
  }
  
  ## CCD
  else if (source == 'ccd') {
    data9 <- get_education_data(
      level = "schools",
      source = "ccd",
      topic = "enrollment",
      filters = list(year = latest_year, grade = 9)) %>% filter(fips==16)
    
    data10 <- get_education_data(
      level = "schools",
      source = "ccd",
      topic = "enrollment",
      filters = list(year = latest_year, grade = 10)) %>% filter(fips==16)
    
    data11 <- get_education_data(
      level = "schools",
      source = "ccd",
      topic = "enrollment",
      filters = list(year = latest_year, grade = 11)) %>% filter(fips==16)
    
    data12 <- get_education_data(
      level = "schools",
      source = "ccd",
      topic = "enrollment",
      filters = list(year = latest_year, grade = 12)) %>% filter(fips==16)
    
    data <- rbind(data9, data10, data11, data12)
  }
  
  ## Bad Input
  else {
    print('ERROR: Unrecognized Source!!!')
    data <- -1
  }
  
  return(data)
}


encode_category <- function(data, categories, vals, target_column_name) {
  # Encode math and reading scores for each category
  generated_columns <- c()
  for (i in 1:length(vals)) {
    val <- vals[i]
    ### Math Test
    math_col_name <- paste0(categories[i], '_math')
    data <- data %>%
      mutate(!!sym(math_col_name) := ifelse(data[[target_column_name]]==val, data$math_test_pct_prof_midpt, 0))
    ### Reading Test
    read_col_name <- paste0(categories[i], '_read')
    data <- data %>%
      mutate(!!sym(read_col_name) := ifelse(data[[target_column_name]]==val, data$read_test_pct_prof_midpt, 0))
    ### Record new category names
    generated_columns <- append(generated_columns, math_col_name)
    generated_columns <- append(generated_columns, read_col_name)
  }
  
  # Aggregate into one record per school
  data <- data %>%
    group_by(ncessch_num) %>%
    select(generated_columns) %>%
    summarize_each(funs(sum))
  
  return(data)
}
```


```{r load_data}
# Load data

###############################################################################

# Pull general data
data <- fetch_idaho_school_data('edfacts', list())

###############################################################################

# Pull and merge race data
### Fetch the data
data_race <- fetch_idaho_school_data('edfacts', list('race'))
### Do something similar to a one-hot encoding
race_categories <- c('white', 'black', 'hispanic', 'asian', 'indian',
                     'hawaiian', 'two_plus', 'alien', 'unknown', 'other',
                     'total')
race_vals <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 20, 99)
data_race_coded <- encode_category(data_race, race_categories, race_vals, 'race')
### Merge
data <- left_join(data, data_race_coded, by='ncessch_num')

###############################################################################

# Pull and merge sex data
### Fetch the data
data_sex <- fetch_idaho_school_data('edfacts', list('sex'))
### Do something similar to a one-hot encoding
sex_categories <- c('male', 'female', 'any')
sex_vals <- c(1, 2, 99)
data_sex_coded <- encode_category(data_sex, sex_categories, sex_vals, 'sex')
### Merge
data <- left_join(data, data_sex_coded, by='ncessch_num')

###############################################################################

# Pull and merge enrollment data
### Fetch the data
data_enrollment <- fetch_idaho_school_data('ccd', list())
### Filter
data_enrollment_staged <- data_enrollment %>%
  select(ncessch_num, enrollment) %>%
  group_by(ncessch_num) %>%
  summarize(enrollment=sum(enrollment))
### Merge
data <- left_join(data, data_enrollment_staged, by='ncessch_num')

###############################################################################

# Pull and merge teacher employment data
## Fetch the data
data_teachers <- get_education_data(level = "schools",
    source = "crdc",
    topic = "teachers-staff",
    filters = list(year = 2017))
### Filter
data_teachers_staged <- data_teachers %>%
  filter(fips==16) %>%
  mutate(ncessch_num=as.numeric(ncessch)) %>%
  select(ncessch_num, teachers_fte_crdc)
### Merge
data <- left_join(data, data_teachers_staged, by='ncessch_num')
### Generate new feature
data$students_per_teacher <- as.numeric(data$enrollment) / as.numeric(data$teachers_fte_crdc)
```

```{r clean_data}
# Convert negative values into NA's
##
## Note: Negative values mean that the data is either missing or suppressed
##       to prevent identifying individual students in small populations
##

formatted_data <- data %>% replace_with_na_all(condition = ~.x <= 0) 
formatted_data_race <- data_race %>% replace_with_na_all(condition = ~.x <= 0) 
formatted_data_sex <- data_sex %>% replace_with_na_all(condition = ~.x <= 0) 

na_count_start <- table(is.na(data))
na_count_end <- table(is.na(formatted_data))

print('NA Values Before Cleaning:')
print(na_count_start)

print('NA Values After Cleaning:')
print(na_count_end)

```

```{r visualize_data}
### Prepare data
melted_formatted_data <- melt(select(formatted_data,
                                     c(read_test_pct_prof_midpt,
                                       math_test_pct_prof_midpt)))

melted_formatted_data_sex=melt(select(formatted_data,
                                      c(male_read,
                                       female_read,
                                       male_math,
                                       female_math)))
melted_formatted_data_sex$sex <- ifelse(melted_formatted_data_sex$variable=='female_math' |
                                        melted_formatted_data_sex$variable=='female_read', 2, 1)
melted_formatted_data_sex$sex <- as.factor(melted_formatted_data_sex$sex)

melted_formatted_data_race=melt(select(formatted_data,
                                       c(white_read,
                                         hispanic_read,
                                         white_math,
                                         hispanic_math,)))
melted_formatted_data_race$race <- ifelse(melted_formatted_data_race$variable=='white_math' |
                                        melted_formatted_data_race$variable=='white_read', 1, 3)
melted_formatted_data_race$race <- as.factor(melted_formatted_data_race$race)

### Boxplot of Total Data
ggplot(data=melted_formatted_data, aes(x=variable, y=value)) +
  geom_boxplot() +
  ggtitle("Box Plot of Test Score Ranges") + 
  labs(subtitle='Idaho High Schools 2017', x='Test Score Range', y='Score')
ggsave(filename="boxplot_all.png",
         width=6,
         height=4)

### Violinplot of Total Data
ggplot(data=melted_formatted_data, aes(x=variable, y=value)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  ggtitle("Violin Plot of Variables")
ggsave(filename="violinplot_all.png",
         width=6,
         height=4)

### Boxplot of Data by Sex
ggplot(aes(x=variable, y=value, fill=sex),
   data=melted_formatted_data_sex) +
   geom_boxplot() +
   scale_fill_manual(labels = c("Male", "Female"),
                     values = c("steelblue4", "darkorange2")) +
   ggtitle("Box Plot of Test Score Ranges (by Sex)") +
   labs(y='Score', x='Test Score Range', subtitle='Idaho High Schools 2017')
ggsave(filename="boxplot_sex.png",
         width=6,
         height=4)

### Boxplot of Data by Race
ggplot(aes(x=variable, y=value, fill=race),
   data=melted_formatted_data_race) +
   geom_boxplot() +
   scale_fill_manual(labels = c("White", "Hispanic"),
                     values = c("mediumslateblue", "orangered3")) +
   ggtitle("Box Plot of Test Score Ranges (by Race)") +
   labs(y='Score', x='Test Score Range', subtitle='Idaho High Schools 2017')
ggsave(filename="boxplot_race.png",
         width=6,
         height=4)
dev.off()
```


```{r visualize_data_scatters}
### Scatterplot of Math Midpt vs. Reading Midpt
ggplot(data=formatted_data, aes(x=math_test_pct_prof_midpt,
                                y=read_test_pct_prof_midpt)) +
  geom_point() +
  labs(x='% of Students Proficient in Math',
       y='% of Students Proficient in Reading',
       title='Math vs. Reading Proficiency',
       subtitle='Idaho High Schools 2017')
ggsave(filename="scatter_all.png",
         width=6,
         height=4)

### Scatterplot of Math Midpt vs. Reading Midpt by Enrollment
ggplot(data=formatted_data, aes(x=math_test_pct_prof_midpt,
                                y=read_test_pct_prof_midpt,
                                size=enrollment)) +
  geom_point(alpha = 1/3) +
  labs(x='% of Students Proficient in Math',
       y='% of Students Proficient in Reading',
       title='Math vs. Reading Proficiency (by Enrollment)',
       subtitle='Idaho High Schools 2017')
ggsave(filename="scatter_enroll.png",
         width=6,
         height=4)

### Scatterplot of Math Midpt vs. Reading Midpt by FTE Teachers
ggplot(data=formatted_data, aes(x=math_test_pct_prof_midpt,
                                y=read_test_pct_prof_midpt,
                                size=students_per_teacher)) +
  geom_point(alpha = 1/3) +
  labs(x='% of Students Proficient in Math',
       y='% of Students Proficient in Reading',
       title='Math vs. Reading Proficiency (by FTE Teachers)',
       subtitle='Idaho High Schools 2017')
ggsave(filename="scatter_student_teacher.png",
         width=6,
         height=4)
dev.off()
```



```{r regressions}
# Regressions
formatted_data_race_sub <- formatted_data_race %>% filter(race != 99)
formatted_data_sex_sub <- formatted_data_sex %>% filter(sex != 99)


### Math Regressions
sex_diff_model_m <- lm(math_test_pct_prof_midpt ~ as.factor(sex),
                     data=formatted_data_sex_sub)
race_diff_model_m <- lm(math_test_pct_prof_midpt ~ as.factor(race),
                      data=formatted_data_race_sub)
size_diff_model_m <- lm(math_test_pct_prof_midpt ~ enrollment,
                     data=data)
fte_diff_model_m <- lm(math_test_pct_prof_midpt ~ students_per_teacher,
                     data=data)


### Reading Regressions
sex_diff_model_r <- lm(read_test_pct_prof_midpt ~ as.factor(sex),
                     data=formatted_data_sex_sub)
race_diff_model_r <- lm(read_test_pct_prof_midpt ~ as.factor(race),
                      data=formatted_data_race_sub)
size_diff_model_r <- lm(math_test_pct_prof_midpt ~ enrollment,
                     data=data)
fte_diff_model_r <- lm(math_test_pct_prof_midpt ~ students_per_teacher,
                     data=data)
```


```{r regressions_summaries}
### Reporting
summary(sex_diff_model_m)
print('##################')
summary(sex_diff_model_r)
```

```{r regressions_summaries}
### Reporting
summary(race_diff_model_m)
print('##################')
summary(race_diff_model_r)
```

```{r regressions_summaries}
### Reporting
summary(size_diff_model_m)
print('##################')
summary(size_diff_model_r)
```

```{r regressions_summaries}
### Reporting
summary(fte_diff_model_m)
print('##################')
summary(fte_diff_model_r)
```


```{r cluster_profiling}
# Cluster schools into categories
## NOTE: Because this is an unsupervised exploratory task, train/test split unnecessary

set.seed(777)
## Subset features
feature_subset_data <- formatted_data %>% select(school_name,
                                                 read_test_pct_prof_midpt,
                                                 math_test_pct_prof_midpt,
                                                 enrollment,
                                                 students_per_teacher)

## Remove rows with NaN (9)
feature_subset_data <- na.omit(feature_subset_data)

##
feature_subset_data_no_name <- feature_subset_data %>% select(-school_name)

## Find a decent K value
elbow_points <- c()
for (k in 3:9) {
  k_model <- kmeans(feature_subset_data_no_name, k)
  elbow_points <- c(elbow_points, k_model$tot.withinss)
}
plot(elbow_points, type='b')
ggsave(filename="elbow.png",
         width=6,
         height=4)
dev.off()

```




```{r cluster}
set.seed(777)
## Build a k_model w/ decent K value
k_model <- kmeans(feature_subset_data_no_name, 5)

feature_subset_data_w_clusters <- feature_subset_data
feature_subset_data_w_clusters$cluster <- as.factor(k_model$cluster)

### Scatterplot of Math Midpt vs. Reading Midpt w/ Cluster
ggplot(data=feature_subset_data_w_clusters,
       aes(x=math_test_pct_prof_midpt,
           y=read_test_pct_prof_midpt,
           color=as.factor(cluster))) +
  geom_point() +
  labs(x='% of Students Proficient in Math',
       y='% of Students Proficient in Reading',
       title='Math vs. Reading Proficiency',
       subtitle='Idaho High Schools 2017')
ggsave(filename="scatter_cluster.png",
         width=6,
         height=4)
dev.off()

### Scatterplot of Math Midpt vs. Reading Midpt w/ Cluster & Enroll
ggplot(data=feature_subset_data_w_clusters,
       aes(x=math_test_pct_prof_midpt,
           y=read_test_pct_prof_midpt,
           color=as.factor(cluster),
           size=enrollment)) +
  geom_point(alpha=7/10) +
  labs(x='% of Students Proficient in Math',
       y='% of Students Proficient in Reading',
       title='Math vs. Reading Proficiency',
       subtitle='Idaho High Schools 2017')
ggsave(filename="scatter_cluster_enroll.png",
         width=6,
         height=4)
dev.off()

## Rough descriptions of each cluster
for (i in 1:5) {
  print('################')
  print(i)
  print(summary(feature_subset_data_w_clusters %>% filter(feature_subset_data_w_clusters$cluster==i)))
}

```



```{r classify_clusters}
feature_subset_data_w_clusters_no_name <- feature_subset_data_w_clusters %>% select(-school_name)
tree_model <- rpart(cluster ~., data=feature_subset_data_w_clusters_no_name)
rpart.plot(tree_model)
png(filename="tree_cluster.png")
dev.off()
```

```{r recursive_clustering}
set.seed(777)
regen_data <- data.frame()
for (i in (1:5)) {
  print(i)
  ### Subset by cluster
  sub_cluster <- feature_subset_data_w_clusters %>% filter(cluster==i)
  
  ### Scale cluster
  sub_cluster_scaled <- sub_cluster %>%
    select(-cluster) %>%
    select(-school_name) %>%
    sapply(rescale, to=c(0:1))

  k_model <- kmeans(sub_cluster_scaled, 4)

  sub_cluster_w_labels <- sub_cluster
  sub_cluster_w_labels$sub_cluster <- as.factor(k_model$cluster)

  ### Scatterplot
  ggplot(data=sub_cluster_w_labels,
         aes(x=math_test_pct_prof_midpt,
             y=read_test_pct_prof_midpt,
             color=as.factor(sub_cluster),
             size=enrollment)) +
    geom_point(alpha = 7/10) +
    labs(x='% of Students Proficient in Math',
         y='% of Students Proficient in Reading',
         title='Math vs. Reading Proficiency',
         subtitle='Idaho High Schools 2017')
  ggsave(filename=paste(i, "sub_cluster_labeled.png", sep="_"),
         width=6,
         height=4)
  
  
  regen_data <- rbind(regen_data, sub_cluster_w_labels)


}

for (i in (1:5)) {
  tree_model <- rpart(cluster ~.,
                      data=data.frame(sub_cluster_list[i])%>%select(-school_name))
  rpart.plot(tree_model)
}

```

```{r inner_cluster_highlight}

critical_schools <- data.frame()
critical_schools <- rbind(critical_schools, regen_data %>% filter(cluster==1) %>% filter(sub_cluster==1) )
critical_schools <- rbind(critical_schools, regen_data %>% filter(cluster==2) %>% filter(sub_cluster==4) )
critical_schools <- rbind(critical_schools, regen_data %>% filter(cluster==3) %>% filter(sub_cluster==3) )
critical_schools <- rbind(critical_schools, regen_data %>% filter(cluster==4) %>% filter(sub_cluster==4) )
critical_schools <- rbind(critical_schools, regen_data %>% filter(cluster==5) %>% filter(sub_cluster==1) )
print(critical_schools)

ggplot(data=critical_schools,
         aes(x=math_test_pct_prof_midpt,
             y=read_test_pct_prof_midpt,
             color=as.factor(cluster),
             size=enrollment)) +
    geom_jitter(alpha = 7/10) +
    labs(x='% of Students Proficient in Math',
         y='% of Students Proficient in Reading',
         title='Math vs. Reading Proficiency',
         subtitle='Idaho High Schools 2017')
  ggsave(filename="critical.png",
         width=6,
         height=4)
  dev.off()
  
write.csv2(critical_schools, file="critical_schools.csv")
```